General constraints on magic distillation protocols from statistical mechanics

### 2021-04-21 magic_paper.tex
small corrections plus references

Our analysis relies on the discrete Wigner representation of quantum states, in which all states and operations can be described on a discrete phase space~\cite{Ferrie_2008, Okay_2021}. Crucially, we focus on magic states with negativity in their Wigner representation, which is known to be a necessary condition for universality in the state-injection model~\cite{cit:veitch, cit:mari, cit:gottesman, cit:knill, Campbell_2011}. Interestingly, taking a thermodynamic perspective in this context raises a problem, namely it is impossible to define entropies for quasi-distributions. We circumvent this obstacle by making use of a more fundamental tool in thermodynamics -- majorization theory \ddd{[cite]}. To our knowledge majorization of quasi-distributions has not been considered in quantum physics before, and therefore our analysis is of broader interest to the study of non-classicality in quantum systems.

@@

Majorisation~\cite{cit:marshall} is a collection of powerful tools that has recently found many applications in quantum information theory~\cite{Nielsen_1999, cit:cwiklinski, cit:lostaglio2, cit:gour, cit:gour2, Horodecki_2003, Vallejos_2021}.
It describes the disorder of distributions that undergo stochastic transformations. In its simplest form it defines a pre-order on probability distributions. Given two distributions $\p= (p_1, \dots, p_n)$ and $\q = (q_1, \dots, q_n)$ over $n$ outcomes, we say that $\p$ majorises $\q$, denoted $\p \succ \q$, if there exists a bistochastic map $A = (A_{ij})$ such that $A\p = \q$, where bistochastic means that $A_{ij} \geq 0$ and $\sum_i A_{ij} = \sum_j A_{ij} = 1$. A well-known result~\cite{cit:marshall} tells us that the condition $ \p \succ \q$ over probability distributions is equivalent to $n-1$ inequalities, which can be checked efficiently.

There is a natural generalisation, which is called $\bmd$--majorization~\cite{Veinott_1971}, or in the context of thermodynamics, thermo-majorization \ddd{[cite]}. For a fixed probability distribution $\r = (r_1, \dots, r_n)$ with positive components, we define majorization relative to $\r$ as $\p \succ_{\r} \q$, if and only if there exists a stochastic map $A$ such that $A\r = \r$ and $A \p = \q$. In order words, the relative majorization pre-order coincides with the image of $\p$ under stochastic maps that have $\r$ as a fixed point. It is readily checked that the original majorization condition between probability distributions corresponds to the case $\r = (1/n, \dots, 1/n)$. The pre-order that results from relative majorization is also easily checked and again corresponds to checking a small number of inequalities.

We can further generalise this to \emph{relative majorization}\ddd{[cite]}, defined as an ordering between pairs of vectors so that 
\begin{equation}
	(\p,\r) \succ (\q, \r')
\end{equation}
iff there is a stochastic map $A$ such that $A\r = \r'$ and $A\p = \q$. We retrieve $\bmd$--majorization when $\r = \r'$.

@@

Central to the analysis is the notion of a Lorenz curve of a vector $\w \in \mathbb{R}^n$ relative to some other vector $\r \in \mathbb{R}^n$. Given a vector $\w$ we define $\w^\downarrow$ to be the re-arrangement of the components of $\w$ into decreasing order. Now, given two $n$--component vectors $\w$ and $\r$ we define the \emph{Lorenz curve} of $\w$ relative to $\r$, denoted $L_{\w|\r}(x)$, as the piece-wise linear function that passes through $(0,0)$ and the $n$ points
\begin{equation}
\label{eq:lc}
        (x_k,L_{\w|\r}(x_k)) =\left( \frac{1}{R}\sum_{i=1}^k r_{\pi(i)}, \sum_{i=1}^k w_{\pi(i)} \right),
\end{equation}
where $R:= \sum_{i=1}^n r_i$, and $\pi$ is the permutation on $n$ objects that sends $\widetilde{\w}$ to $\widetilde{\w}^{\downarrow}$, where $\widetilde{w}_i := w_i/r_i$ is the vector of component-wise ratios of $\w$ and $\r$. The form of this requires that $\r$ has no zero components, which we shall assume without loss of generality as in any physical situation the rank of a quantum state is not an operationally meaningful quantity.
We also observe that the curve contains non-differentiable points, called \emph{elbows} at locations where $\widetilde{\bmw}$ changes value.

If $\w$ and $\r$ are both probability distributions then the Lorenz curve is defined on the interval $[0,1]$, and rises monotonically until it reaches the value $1$ at $x=1$. The value $L_{\w|\r}(x) = 1$ is a global maximum, attained at the end-point $x=1$. Moreover, if $\w, \w', \r, \r'$ are all valid probability distributions with neither of $\r,\r'$ having negative components then it has been shownin~\cite{ruch_mixing_1978} that
\begin{equation}
(\w, \r) \succ (\w', \r') \mbox{ if and only if } L_{\w |\r}(x) \ge L_{\w' |\r'}(x).
\end{equation}
This provides a simple way of computing whether relative majorization holds between pairs of probability distributions.

@@

However, relative majorization is usually considered for probability distributions, and so we must verify that the same Lorenz curve conditions apply to quasi-probability distributions. This can be done from first principles, but a simpler way is see that this holds is to reduce to the problem of genuine distributions by first `masking' the negativity in the quasi-probability distribution $\w$ using the reference distribution $\r$ and then applying the conditions for relative majorization.

More precisely, if the components of $\r$ are never zero, then there always exists an $\epsilon >0$ such that $\w_\epsilon := \epsilon \w + (1-\epsilon) \r$ is a genuine probability distribution. We can see that $(\w , \r) \succ (\w', \r')$ if and only if $(\w_\epsilon , \r) \succ (\w_\epsilon', \r')$, where we choose $\epsilon$ so that both $\w_\epsilon$ and $\w_\epsilon'$ are probability distributions. This equivalence holds because there exists a stochastic map $A$ such that $A \w = \w'$ and $A \r = \r'$ if and only if 
\begin{equation}
A[\epsilon \w + (1-\epsilon) \r] = \epsilon \w' + (1-\epsilon) \r'\mbox{ and } A\r = \r'.
\end{equation}
As mentioned, relative majorization between pairs of distributions occurs if and only if the associated Lorenz curves never cross. Therefore we can express this condition in terms of the Lorenz curves of the quasi-probability distributions $\w$ and $\w'$ relative to $\r$ and $\r'$ respectively.

@@

As discussed already, the free operations within the magic theory $\R$ are represented by stochastic maps, and within $\R_\sigma$ by stochastic maps that leave $W_\sigma(\x)$ invariant. Therefore, a necessary condition for magic state transformations $\rho_1 \rightarrow \rho_2$ within $\R_\sigma$ will be that 
\begin{equation}
W_{\rho_1} \succ_{W_{\sigma}} W_{\rho_2},
\end{equation}
or put another way, relative to $W_\sigma$ the quasi-distribution $W_{\rho_1}$ is more ordered than the quasi-distribution $W_{\rho_2}$.

This in turn can be expressed in terms of Lorenz curves. To simplify notation we denote by $L_{\rho | \sigma}(x)$ the Lorenz curve $L_{W_{\rho} | W_{\sigma}} (x)$. We therefore have that
\begin{equation}
\rho_1 \rightarrow \rho_2 \mbox{ within }\R_\sigma \mbox{ implies } L_{\rho_1 |\sigma} (x) \ge L_{\rho_2 |\sigma} (x),
\end{equation}
for all $x \in [0,1]$. Therefore, the magic state Lorenz curve condition can be used to impose upper bounds on magic distillation rates. Note, however, that unlike any monotone, this is not a single numerical constraint, but a \emph{family} of constraints. Indeed, for $n$ copies of a $d$--dimensional system the number of terms in $W_{\rho}$ is $d^{2n}$, so the Lorenz curve condition corresponds to \emph{exponentially} many constraints, of which there are as many independent constraints as there are elbows in the Lorenz curve of the output state, which is proved in~\cref{thm:elbows} in~\cref{app:frag}.
In principle, any one location $x$ provides a valid constraint leading to some upper distillation bound and optimising over the location provides the strictest bound.

@@

In this section we give some generic aspects of Lorenz curves for magic states. In particular these allow us to interpret previous magic monotones in a new light. The f,irst result provides an extremely simple way to see that the sum-negativity of a magic state is a monotone.

@@

If we have a magic state $\rho$ that has negativity in its Wigner representation, then as discussed, its Lorenz curve over-shoots $1$ and reaches a non-trivial maximum $L_\star$ that depends on the particular state. There is an exact relation between the peak and sum-negativity or mana, which is provided by the following result. 
\begin{lemma}\label{lem:lcmax}
	Given a magic state $\rho$, the maximum $L_\star$ of its Lorenz curve $\lc{\rho}{\sigma}(x)$ is independent of the $\sigma$--fragment and equal to $1+\sn{\rho}$. Moreover, the majorization constraint is stronger than mana in every fragment.
\end{lemma}
The proof of this is given in Appendix \ref{lem:lcmax}. Using the Lorenz curve perspective, it is simple to construct other monotones. For example, within a given fragment, the area above $L=1$ is another magic monotone.
\begin{lemma} Let $\A_\sigma(\rho)$ be the area of the region $\{(x,y) : (x, L_{\rho | \sigma}(x)) \ge 1\}$. Then $\A_\sigma(\rho)$ is a magic monotone for $\R_\sigma$.
\end{lemma}
\begin{proof}
Consider the transformation $\rho_1 \rightarrow \rho_2$ within $\R_\sigma$. We have that $L_{\rho_1|\sigma}(x)$ is never below the curve $L_{\rho_2|\sigma}(x)$, and therefore the region above $L=1$ for $\rho_2$ is a subset of the corresponding region for $\rho_1$. Thus, $\A_\sigma(\rho_1) \geq \A_\sigma(\rho_2)$, so $\A_\sigma$ is a magic monotone in $\R_\sigma$.
\end{proof}
In contrast to mana, the area monotone is \emph{specific to the fragment}, and its value will vary as we change $\sigma$. Therefore its monotonicity is also including the physics of the fragment and provides a means to analyse magic distillation only under free operations that leave $\sigma$ invariant.

We ultimately want to make statements about asymptotic distillation rates $R$,
\begin{equation}
\rho_1^{\otimes N} \longrightarrow \rho_2^{\otimes R N}
\end{equation}
where $\rho_2$ is a purer magic state than $\rho_1$ and $R$ is made as large as possible as $N$ tends to infinity. We shall next turn to this analysis, and exploit basic features of a magic state in order to obtain non-trivial distillation bounds.

@@

This means in particular that $\E(\tau^{\otimes n}) = \tau'$ is a symmetric state on $m$ subsystems, and so by the quantum de Finetti theorem \ddd{[cite]} we have that
\begin{equation}
\tau'  = \int d\mu(x) \sigma_x^{\otimes m},
\end{equation}
where $d\mu(x)$ is a probability measure over a set $\{\sigma_x\}$ of single qudit states.

@@

  For example, suppose that $H=H'$ then this implies the protocol would leave the equilibrium state unchanged and so corresponds to a Gibbs-preserving map \ddd{[cite Philippe]}. In the more general case $\Delta F := F' - F \ne 0$ and we have a free energy difference between the initial and final times, with respect to the relevant Hamiltonians. This is similar to the free energy differences that occur in fluctuation relations, where the final state of the system can be arbitrarily far from equilibrium, yet a free energy term for the final system still arises.
Firstly, the specific numerical factors in $\epsilon_\star$ are a result of our choice of magic state. 
Secondly, these bounds are derived based on analysis of only a part of the Lorenz curves and can be improved via a finer analysis. 
This is apparent by simple numerical calculations on the entirety of the curves.
Specifically, the existing bounds follow by considering the dominant terms in the rescaled Wigner distribution and do not, for example, take into account the Lorenz curve's peak structure.

>>>>

Our analysis relies on the discrete Wigner representation of quantum states, in which all states and operations can be described on a discrete phase space~\cite{Ferrie_2008, Okay_2021}. Crucially, we focus on magic states with negativity in their Wigner representation, which is known to be a necessary condition for universality in the state-injection model~\cite{cit:veitch, cit:mari, cit:gottesman, cit:knill, Campbell_2011}. Interestingly, taking a thermodynamic perspective in this context raises a problem, namely it is impossible to define entropies for quasi-distributions. We circumvent this obstacle by making use of a more fundamental tool in thermodynamics -- majorization theory~\cite{cit:marshall, Veinott_1971, Ruch_1976}. To our knowledge majorization of quasi-distributions has not been considered in quantum physics before, and therefore our analysis is of broader interest to the study of non-classicality in quantum systems.

@@

Majorisation is a collection of powerful tools that has recently found many applications in quantum information theory~\cite{Nielsen_1999, cit:cwiklinski, cit:lostaglio2, cit:gour, cit:gour2, Horodecki_2003, Vallejos_2021}.
It describes the disorder of distributions that undergo stochastic transformations. In its simplest form it defines a pre-order on probability distributions. Given two distributions $\p= (p_1, \dots, p_n)$ and $\q = (q_1, \dots, q_n)$ over $n$ outcomes, we say that $\p$ majorises $\q$, denoted $\p \succ \q$, if there exists a bistochastic map $A = (A_{ij})$ such that $A\p = \q$, where bistochastic means that $A_{ij} \geq 0$ and $\sum_i A_{ij} = \sum_j A_{ij} = 1$. A well-known result~\cite{cit:marshall} tells us that the condition $ \p \succ \q$ over probability distributions is equivalent to $n-1$ inequalities, which can be checked efficiently.

There is a natural generalisation, which is called $\bmd$--majorization~\cite{Veinott_1971}, or in the context of thermodynamics, thermo-majorization~\cite{cit:horodecki2013}. For a fixed probability distribution $\r = (r_1, \dots, r_n)$ with positive components, we define majorization relative to $\r$ as $\p \succ_{\r} \q$, if and only if there exists a stochastic map $A$ such that $A\r = \r$ and $A \p = \q$. In order words, this majorization pre-order coincides with the image of $\p$ under stochastic maps that have $\r$ as a fixed point. It is readily checked that the original majorization condition between probability distributions corresponds to the case $\r = (1/n, \dots, 1/n)$. The pre-order that results from relative majorization is also easily checked and again corresponds to checking a small number of inequalities.

We can further generalise this to \emph{relative majorization}~\cite{Ruch_1976, ruch_mixing_1978, Renes_2016, Buscemi_2017}, defined as an ordering between pairs of vectors so that 
\begin{equation}
	(\p,\r) \succ (\q, \r')
\end{equation}
iff there is a stochastic map $A$ such that $A\r = \r'$ and $A\p = \q$. We retrieve $\bmd$--majorization when $\r = \r'$.

@@

Central to the analysis is the notion of a Lorenz curve of a vector $\w \in \mathbb{R}^n$ relative to some other vector $\r \in \mathbb{R}^n$. Given a vector $\w$ we define $\w^\downarrow$ to be the re-arrangement of the components of $\w$ into decreasing order. Given two $n$--component vectors $\w$ and $\r$, we first define $\widetilde{\w}$, where $\widetilde{w}_i := w_i/r_i$, as the vector of component-wise ratios between $\w$ and $\r$.
We can now define the \emph{Lorenz curve} of $\w$ relative to $\r$, denoted $L_{\w|\r}(x)$, as the piece-wise linear function that passes through $(0,0)$ and the $n$ points
\begin{equation}
\label{eq:lc}
        (x_k,L_{\w|\r}(x_k)) =\left( \frac{1}{R}\sum_{i=1}^k r_{\pi(i)}, \sum_{i=1}^k w_{\pi(i)} \right),
\end{equation}
where $R:= \sum_{i=1}^n r_i$ and $\pi$ is the permutation on $n$ objects mapping $\widetilde{\w}$ to $\widetilde{\w}^{\downarrow}$. The form of this requires that $\r$ has no zero components, which we shall assume without loss of generality as in any physical situation the rank of a quantum state is not an operationally meaningful quantity.
We also observe that the curve contains non-differentiable points, called \emph{elbows} at locations where $\widetilde{\bmw}$ changes value.

If $\w$ and $\r$ are both probability distributions then the Lorenz curve is defined on the interval $[0,1]$, and rises monotonically until it reaches the value $1$ at $x=1$. The value $L_{\w|\r}(x) = 1$ is a global maximum, attained at the end-point $x=1$. Moreover, if $\w, \w', \r, \r'$ are all valid probability distributions with neither of $\r,\r'$ having negative components then it has been shown in~\cite{ruch_mixing_1978} that
\begin{equation}
(\w, \r) \succ (\w', \r') \mbox{ if and only if } L_{\w |\r}(x) \ge L_{\w' |\r'}(x).
\end{equation}
This provides a simple way of computing whether relative majorization holds between pairs of probability distributions.

@@

However, relative majorization is usually considered for probability distributions, so we need to verify that the same Lorenz curve conditions apply to quasi-probability distributions before proceeding with our analysis. This can be done from first principles, but a simpler way is see that this holds is to reduce to the problem of genuine distributions by first `masking' the negativity in the quasi-probability distribution $\w$ using the reference distribution $\r$ and then applying the conditions for relative majorization.

More precisely, since the components of $\r$ are strictly positive, there always exists an $\epsilon >0$ such that $\w_\epsilon := \epsilon \w + (1-\epsilon) \r$ is a genuine probability distribution. We can see that $(\w , \r) \succ (\w', \r')$ if and only if $(\w_\epsilon , \r) \succ (\w_\epsilon', \r')$, where we choose $\epsilon$ so that both $\w_\epsilon$ and $\w_\epsilon'$ are probability distributions. This equivalence holds because there exists a stochastic map $A$ such that $A \w = \w'$ and $A \r = \r'$ if and only if 
\begin{equation}
A[\epsilon \w + (1-\epsilon) \r] = \epsilon \w' + (1-\epsilon) \r'\mbox{ and } A\r = \r'.
\end{equation}
As mentioned, relative majorization between pairs of distributions occurs if and only if the associated Lorenz curves never cross. Therefore, we can express this condition in terms of the Lorenz curves of the quasi-probability distributions $\w$ and $\w'$ relative to $\r$ and $\r'$ respectively.

@@

As discussed already, the free operations within the magic theory $\R$ are represented by stochastic maps, and within $\R_\sigma$ by stochastic maps that leave $W_\sigma(\x)$ invariant. Therefore, a necessary condition for magic state transformations $\rho_1 \rightarrow \rho_2$ within $\R_\sigma$ will be that 
\begin{equation}
	W_{\rho_1} \succ_{W_{\sigma}} W_{\rho_2},
\end{equation}
or put another way, that the quasi-distribution $W_{\rho_1}$ is more ordered than the quasi-distribution $W_{\rho_2}$ relative to $W_\sigma$.

This in turn can be expressed in terms of Lorenz curves. To simplify notation, we denote by $L_{\rho | \sigma}(x)$ the Lorenz curve $L_{W_{\rho} | W_{\sigma}} (x)$. We therefore have that
\begin{equation}
\rho_1 \rightarrow \rho_2 \mbox{ within }\R_\sigma \mbox{ implies } L_{\rho_1 |\sigma} (x) \ge L_{\rho_2 |\sigma} (x),
\end{equation}
for all $x \in [0,1]$. Therefore, the magic state Lorenz curve condition can be used to impose upper bounds on magic distillation rates. Note, however, that unlike any monotone, this is not a single numerical constraint, but a \emph{family} of constraints. Indeed, for $n$ copies of a $d$--dimensional system the number of terms in $W_{\rho}$ is $d^{2n}$, so the Lorenz curve condition corresponds to exponentially many constraints, of which though there are at most as many independent constraints as there are elbows in the Lorenz curve of the output state, as proved in~\cref{thm:elbows} in~\cref{app:frag}.
In principle, any one location $x$ provides some valid constraint leading to an upper distillation bound, while optimising over the location provides the strictest majorisation bound.

@@

In this section, we give some generic aspects of Lorenz curves for magic states. In particular, these allow us to interpret previous magic monotones in a new light. The first result provides an extremely simple way to see that the sum-negativity of a magic state is a monotone.

@@

If we have a magic state $\rho$ that has negativity in its Wigner representation, then, as discussed, its Lorenz curve over-shoots $1$ and reaches a non-trivial maximum $L_\star$ that depends on the particular state. There is an exact relation between $L_\star$ and sum-negativity or mana, which is provided by the following result. 
\begin{lemma}\label{lem:lcmax}
	Given a magic state $\rho$, the maximum $L_\star$ of its Lorenz curve $\lc{\rho}{\sigma}(x)$ is independent of the $\sigma$--fragment and equal to $1+\sn{\rho}$. Moreover, the majorization constraint is stronger than mana in every fragment.
\end{lemma}
The proof of this is given in Appendix \ref{lem:lcmax}. Using the Lorenz curve perspective, it is simple to construct other monotones. For example, within a given fragment, the area above $L=1$ is another magic monotone.
\begin{lemma}
Given a magic state $\rho$ and a free state $\sigma$, let $\A_\sigma(\rho)$ be the area of the region $\{(x, y): 1 \leq y \leq L_{\rho | \sigma}(x)\}$. Then $\A_\sigma(\rho)$ is a magic monotone for $\R_\sigma$.
\end{lemma}
\begin{proof}
Consider the transformation $\rho_1 \rightarrow \rho_2$ within $\R_\sigma$. We have that $L_{\rho_1|\sigma}(x)$ is never below the curve $L_{\rho_2|\sigma}(x)$, and therefore the region above $L=1$ for $\rho_2$ is a subset of the corresponding region for $\rho_1$. Thus, $\A_\sigma(\rho_1) \geq \A_\sigma(\rho_2)$, so $\A_\sigma$ is a magic monotone in $\R_\sigma$.
\end{proof}
In contrast to mana, the area monotone is \emph{specific to the fragment}, and its value will vary as we change $\sigma$. Therefore its monotonicity is also including the physics of the fragment and provides a means to analyse magic distillation only under free operations that leave $\sigma$ invariant.

We ultimately want to make statements about asymptotic distillation rates $R$,
\begin{equation}
\rho_1^{\otimes n} \longrightarrow \rho_2^{\otimes R n}
\end{equation}
where $\rho_2$ is a purer magic state than $\rho_1$ and $R$ is made as large as possible as $n$ tends to infinity. We shall next turn to this analysis, and exploit basic features of a magic state in order to obtain non-trivial distillation bounds.

@@

This means in particular that $\E(\tau^{\otimes n}) = \tau'$ is a symmetric state on $m$ subsystems, and so by the quantum de Finetti theorem~\cite{christandl_2007} we have that
\begin{equation}
\tau'  = \int d\mu(x) \sigma_x^{\otimes m},
\end{equation}
where $d\mu(x)$ is a probability measure over a set $\{\sigma_x\}$ of single qudit states.

@@

  For example, suppose that $H=H'$ then this implies the protocol would leave the equilibrium state unchanged and so corresponds to a Gibbs-preserving map~\cite{faist_2015}. In the more general case $\Delta F := F' - F \ne 0$ and we have a free energy difference between the initial and final times, with respect to the relevant Hamiltonians. This is similar to the free energy differences that occur in fluctuation relations, where the final state of the system can be arbitrarily far from equilibrium, yet a free energy term for the final system still arises.
Firstly, the specific numerical factors in $\epsilon_\star$ are a result of our choice of magic state. 
Secondly, these bounds are derived based on analysis of only a part of the Lorenz curves and can be improved via a finer analysis. 
This is apparent by simple numerical calculations on the entirety of the curves.
Specifically, the existing bounds follow by considering the dominant terms in the rescaled Wigner distribution and do not, for example, take into account the Lorenz curve's peak structure.
